---
title: "SDS HW1"
author: "Vlado Vukovic"
date: "11/5/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1
My assumptions are that Rome has population of 1,486,981 people, and one phone number is available per 4 people (avg family) + one phone number per 1,000 people for other stuff like businesses or government. 

Line busyness is a normal random variable with mean 0.2 and sd 0.1.

```{r}
pop = 1486981
totalNumbers = pop / 4 +  pop /1000 
probBusy = rnorm(1,0.2,0.1)
prior = 1 / totalNumbers
post = 1 / (totalNumbers*probBusy)
cat("My belief was", prior, "\n")
cat("My belief after dialing is", post, "\n")
cat("Therefore my belief increased by", post/prior, "times")
```

## Question 2
### Part 1

```{r}
result <- function(game) {
  w <- array(0, dim=c(1,2))
  
  row1 = game[1] + game[2] + game[3]
  if (row1 == 0) {w[1] <- 1} else if (row1 == 3) {w[2] <- 1}
  
  row2 = game[4] + game[5] + game[6]
  if (row2 == 0) {w[1] <- 1} else if (row2 == 3) {w[2] <- 1}
  
  row3 = game[7] + game[8] + game[9]
  if (row3 == 0) {w[1] <- 1} else if (row3 == 3) {w[2] <- 1}
  
  col1 = game[1] + game[4] + game[7]
  if (col1 == 0) {w[1] <- 1} else if (col1 == 3) {w[2] <- 1}
  
  col2 = game[2] + game[5] + game[8]
  if (col2 == 0) {w[1] <- 1} else if (col2 == 3) {w[2] <- 1}
  
  col3 = game[3] + game[6] + game[9]
  if (col3 == 0) {w[1] <- 1} else if (col3 == 3) {w[2] <- 1}
  
  diag1 = game[1] + game[5] + game[9]
  if (diag1 == 0) {w[1] <- 1} else if (diag1 == 3) {w[2] <- 1}
  
  diag2 = game[3] + game[5] + game[7]
  if (diag2 == 0) {w[1] <- 1} else if (diag2 == 3) {w[2] <- 1}
  
  if(w[1] == 0 & w[2] == 0) {return("t")} #reprint("Its a tie")
  if(w[1] == 1 & w[2] == 0) {return("o")} #print("O wins")}
  if(w[1] == 0 & w[2] == 1) {return("x")} #print("X wins")}
  if(w[1] == 1 & w[2] == 1) {return("t")} #print("Its a tie")}
}

res <- array(NA, dim=c(1,1000))
for (k in 1:1000) {
  first = matrix(NA,3,3)
  for (i in 1:9){
    first[i]= rbinom(1,1,0.5)
  }
  res[k] <- result(first)
}
t <- sum(res == "t")
o <- sum(res == "o")
x <- sum(res == "x")
cat("Out of 1,000 games: X won", x, ", O won", o, "and there were", t, "ties")
cat("Therefore probability that X wins is :", (x/1000))
```

### Part 2

```{r}

result1 <- function(game) {
  w <- array(0, dim=c(1,2))
  row1 = game[1] + game[2] + game[3]
  if (row1 == 0) {w[1] <- 1}
  row2 = game[4] + game[5] + game[6]
  if (row2 == 0) {w[1] <- 1}
  row3 = game[7] + game[8] + game[9]
  if (row3 == 0) {w[1] <- 1}
  col1 = game[1] + game[4] + game[7]
  if (col1 == 0) {w[1] <- 1}
  col2 = game[2] + game[5] + game[8]
  if (col2 == 0) {w[1] <- 1}
  col3 = game[3] + game[6] + game[9]
  if (col3 == 0) {w[1] <- 1}
  diag1 = game[1] + game[5] + game[9]
  if (diag1 == 0) {w[1] <- 1}
  diag2 = game[3] + game[5] + game[7]
  if (diag2 == 0) {w[1] <- 1}
  if(w[1] == 1) {return("o")} #print("O wins")}
}

result2 <- function(game) {
  w <- array(0, dim=c(1,2))
  row1 = game[1] + game[2] + game[3]
  if (row1 == 3) {w[2] <- 1}
  row2 = game[4] + game[5] + game[6]
  if (row2 == 3) {w[2] <- 1}
  row3 = game[7] + game[8] + game[9]
  if (row3 == 3) {w[2] <- 1}
  col1 = game[1] + game[4] + game[7]
  if (col1 == 3) {w[2] <- 1}
  col2 = game[2] + game[5] + game[8]
  if (col2 == 3) {w[2] <- 1}
  col3 = game[3] + game[6] + game[9]
  if (col3 == 3) {w[2] <- 1}
  diag1 = game[1] + game[5] + game[9]
  if (diag1 == 3) {w[2] <- 1}
  diag2 = game[3] + game[5] + game[7]
  if (diag2 == 3) {w[2] <- 1}
  if(w[2] == 1) {return("x")} #print("X wins")}
}

res1 <- array(NA, dim=c(1,1000))
for (j in 1:1000) {
  second = matrix(100,3,3)
  allCells <- seq(1:9)
  remaining <- allCells
  for (i in 1:9){
    selected <- sample(remaining,1,replace=FALSE)
    if (i == 9) {
        selected <- remaining
    }
    remaining <- setdiff(remaining,selected)
    if (i %% 2 == 0) {
      second[selected] <- 0
      if (length(result1(second)) > 0) {
        res1[j] <- result1(second) 
        break
      }
    }
    else {
      second[selected] <- 1
      if (length(result2(second)) > 0) {
        res1[j] <- result2(second) 
        break
        }
    }
  }
}
res1[is.na(res1)] <- "t"
t1 <- sum(res1 == "t")
o1 <- sum(res1 == "o")
x1 <- sum(res1 == "x")
cat("Out of 1,000 games: X won", x1, ", O won", o1, "and there were", t1, "ties")
cat("Therefore probability that X wins is :", (x1/1000))
cat("while probability that O wins is :", (o1/1000))
```

## Question 3

Monty Hall works because we initially assign .66 probability to doors that we dont pick, upon switching (opening one of the doors) the probability of .66 remains (on paper) on the remaining one door, hence we should always switch. Which is also confirmed through simulation below.

Therefore it doesnt matter what we initially choose, we should switch in the long run although we loose when we pick door with car which is 0.33.

```{r}
doors <- c("A","B","C")
decision <- c()
for (i in 1:1000) 
{
  prize <- sample(doors)[1]
  pick <- sample(doors)[1]
  open <- sample(doors[which(doors != prize & doors != pick)])[1]
  switch <- doors[which(doors != pick & doors != open)]
  if(pick == prize){decision=c(decision,"NoSwitch")}
  if(switch == prize){decision=c(decision,"Switch")}
}
cat("Switching was appropriate strategy in :", (length(which(decision == "NoSwitch"))/1000), "of times")
cat("Switching was appropriate strategy in :", (length(which(decision == "Switch"))/1000), "of times")
```

## Question 4
### Part 1

```{r Part1}
xx <- runif(100,0,1)
y <- array(NA, dim=c(1,100))
for(i in 1:100) {
  x <- xx[i]
  if (x <= 0.5) {
    x <- 2*x
  }
  else {
    x <- 2*(1-x)
  }
  y[i] <- x
}

x <- xx
par(mfrow=c(2,3))
plot(seq(1:100),x, type = 'l',col="Green", xlab="n", ylab="value", main="Values of Uniform")
hist(x, breaks=10, col="blue", main="Hist of Uniform")
curve(dunif(x),add=T, col="red", lwd=1)
plot(density(x),main="Density of Uniform")

remove(x)
x <- y
plot(seq(1:100),x, type = 'l', col="Green", xlab="n", ylab="value", main="Values of Transform")
hist(x, breaks=10, col="blue", main="Hist of Transform")
curve(dunif(x),add=T, col="red", lwd=1)
plot(density(x), main="Density of Transform")
```

### Part 2
Generate 1 uniform point and generate 99 consective transforms.

The properties are that transformations hit zero upon reaching value of 0.5.

```{r Part2, echo=FALSE}
remove(x,xx,y)

xx <- runif(1,0,1)
y <- array(NA, dim=c(1,100))
y[1] <- xx
for(i in 2:100) {
    x <- y[i-1]
  if (x <= 0.5) {
    x <- 2*x
  }
  else {
    x <- 2*(1-x)
  }
  y[i] <- x
}

par(mfrow=c(1,3))
x <- y
plot(seq(1:100),x, type = 'l', col="Green", xlab="n", ylab="value", main="Values of Transform")
hist(x, breaks=10, col="blue", main="Hist of Transform")
curve(dunif(x),add=T, col="red", lwd=1)
plot(density(x), main="Density of Transform")

```

### Part 3

The resulting sequence has desirible behavior considering that most of values are edge cases.

```{r Part3, echo=FALSE}
remove(x,xx,y)

k <- rnorm(1,5,3)
xx <- 1/2^k
y <- array(NA, dim=c(1,100))
y[1] <- xx
for(i in 2:100) {
  x <- y[i-1]
  if (x <= 0.5) {
    x <- 2*x
  }
  else {
    x <- 2*(1-x)
  }
  y[i] <- x
}

par(mfrow=c(1,3))
x <- y
plot(seq(1:100),x, type = 'l', col="Green", xlab="n", ylab="value", main="Values of Transform")
hist(x, breaks=10, col="blue", main="Hist of Transform")
curve(dunif(x),add=T, col="red", lwd=1)
plot(density(x), main="Density of Transform")

```
